---
title: "Does elliptical biface morphology differ as a function of raw material color?"
author:
  - name: Robert Z. Selden, Jr.
    affiliations:
      - Heritage Research Center, Stephen F. Austin State University; Department of Biology, Stephen F. Austin State University; Texas Archeological Research Laboratory, The University of Texas at Austin; Cultural Heritage Department, Jean Monnet University
    email: zselden@sfasu.edu
    orcid: 0000-0002-1789-8449
bibliography: [book.bib]
date: "`r format(Sys.time(), '%d %B, %Y')`"
link-citations: yes
github-repo: seldenlab/elliptical.bifaces.1
description: Supplementary Materials
---

>This process of comparison, of recognising in one form a definite permutation or _deformation_ of another, apart altogether from a precise and adequate understanding of the original "type" or standard of comparison, lies within the immediate province of mathematics, and finds its solution in the elementary use of a certain method of the mathematician [@RN7522].

# Basis of inquiry 

Archaeologists regularly assume that the physical properties of lithic raw materials are a determining factor in the morphology of stone tools; although, some have argued that this assumption warrants regular testing. Raw material nodule size and mechanical flaking properties are routinely advanced to account for morphological differences that occur between sites and assemblages. Lithic raw material assignments often use color as the determining characteristic, and new analytical tools allow for an objective analysis of raw material color. To evaluate the assumption that elliptical biface shape is a function of raw material color, each elliptical biface from 41AN13 was assigned to a raw material color group, and subsequently analyzed using the tools of geometric morphometrics. Results demonstrate that elliptical bifaces differ significantly in shape, but not size, by raw material color group. This finding supports the interpretation that extralocal producers—and local Caddo knappers—conditioned elliptical biface shape not based upon nodule size or mechanical flaking properties, but on the basis of raw material color.

## Keywords

Caddo; lithics; computational archaeology; archaeoinformatics; museum studies; digital humanities; non-Western art history

# Acknowledgments {-}

My thanks to the Caddo Nation of Oklahoma, the Caddo Nation Tribal Council, Tribal Chairman, and Tribal Historic Preservation Office for permission and access to NAGPRA and previously repatriated collections. I also extend my gratitude to Lauren Bussiere at the Texas Archeological Research Laboratory for her assistance with access to the elliptical bifaces. Thanks also to John Harman for access to the DStretch plugin for ImageJ, to John E. Dockall, David H. Dye, Harry J. Shafer, Hiram F. (Pete) Gregory, Timothy K. Perttula, Christian S. Hoggard, and David K. Thulman for their comments and constructive criticisms on the ongoing analyses of Caddo bifaces, and to Emma Sherratt, Kersten Bergstrom, Dean C. Adams, and Michael L. Collyer for their constructive criticisms, general comments, and suggestions throughout the development of this research program.

# Funding {-}

Components of this analytical work flow were developed and funded by a Preservation Technology and Training grant (P14AP00138) to RZS from the National Center for Preservation Technology and Training (NCPTT), and additional grants to RZS from the Caddo Nation of Oklahoma, National Forests and Grasslands in Texas (15-PA-11081300-033) and the United States Forest Service (20-PA-11081300-074). Funding and logistical support to analyze the bifaces from 41AN13 was provided by the Heritage Research Center at Stephen F. Austin State University.

# Data management {-}

Reproducibility---the ability to recompute results---and replicability---the chances other experimenters will achieve a consistent result---are two foundational characteristics of successful scientific research [@RN20918]. The analysis code associated with this project can be accessed through this document, is available in the GitHub repository ([https://github.com/seldenlab/elliptical.bifaces.1](https://github.com/seldenlab/elliptical.bifaces.1)), and digitally curated on the Open Science Framework ([DOI: 10.17605/OSF.IO/2CHFE](https://osf.io/2chfe/)). The reproducible nature of this enterprise provides a means for others to critically assess and evaluate the various analytical components [@RN20915;@RN20916;@RN20917], which is a necessary requirement for the production of reliable knowledge.

Reproducibility projects in [psychology](https://osf.io/ezcuj/) and [cancer biology](https://www.cos.io/rpcb) are impacting current research practices across all domains. Examples of reproducible research are becoming more abundant in archaeology [@RN20804;@RN21009;@RN11783;@RN21001;@RN9364;@RN11264], and the next generation of archaeologists are learning those tools and methods needed to reproduce and/or replicate research results [@RN21007]. Reproducible and replicable research work flows are often employed at the highest levels of humanities-based inquiries to mitigate concern or doubt regarding proper execution, and is of particular import should the results have---explicitly or implicitly---a major impact on scientific progress [@RN21008].

# Colophon {-}

This version of the analysis was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# what R packages and versions were used?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

Current Git commit details are:

```{r}
# where can I find this commit? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
